---
title: "01_dataset_creation"
author: "Miguel Ángel Armengol & Jay Chandra"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_notebook:
    code_folding: hide
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes

knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = paste0(substr(inputFile,1,nchar(inputFile)-4)," ",Sys.Date(),'.html')) })
---

# Environment

```{r message=FALSE, warning=FALSE}
library(bigrquery)
library(summarytools)
library(readr)
library(stringr)
library(sqldf)
library(dplyr)
library(tableone)
library(Hmisc)
library(caret)
library(plotly)
library(table1)
library(Amelia)
options(scipen=999)

# useful functions
impute.median <- function(x) replace(x, is.na(x), median(x, na.rm = TRUE))
'%!in%' <- function(x,y)!('%in%'(x,y))

```


# Set up BigQuery related functions

This chunks also creates the run_query and get_sql function.

```{r setup, include=FALSE}
# Updated for our year
project_id <- "hst-953-2019"
options(httr_oauth_cache=FALSE)
# Function that takes in a sql command and runs it on bigquery
run_query <- function(query){
  data <- query_exec(query, project=project_id, use_legacy_sql=FALSE,max_pages = Inf)
  return(data)
}

# function for reading sql files
getSQL <- function(filepath){
  con = file(filepath, "r")
  sql.string <- ""

  while (TRUE){
    line <- readLines(con, n = 1)

    if ( length(line) == 0 ){
      break
    }

    line <- gsub("\\t", " ", line)

    if(grepl("--",line) == TRUE){
      line <- paste(sub("--","/*",line),"*/")
    }

    sql.string <- paste(sql.string, line)
  }

  close(con)
  return(sql.string)
}

'%!in%' <- function(x,y)!('%in%'(x,y))
```

# Loading queries and extracting the data

Loads all queries from the sql files in the extraction folder and runs them into RBigQuey to extract the data.

```{r}
hct <- run_query(getSQL("sql/hct.sql" ))
# removes missing values

# patients excluded due to output amount
IO_exclusion <- run_query(getSQL('sql/IO_exclusion.sql'))

# comorbidities
elixhauser_score <- run_query(getSQL('sql/elixhauser_score.sql'))

# severity score
oasis_related <- run_query(getSQL('sql/oasis_related.sql'))

# SOFA related variables
sofa_total_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_total_day1_to_day4.sql"))

# List of patients that are bleeding we want to exclude.
patient_inexcluded_icd9 <- run_query(getSQL('sql/patient_inexcluded_icd9.sql'))

# Sepsis patients according to angus criteria
sepsis <- run_query(getSQL('sql/sepsis.sql'))

# intake data
intake <- run_query(getSQL("sql/intake.sql"))
intake_full <- run_query(getSQL("sql/intake_full.sql"))
intake72hrs <- run_query(getSQL("sql/intake_72hrs.sql"))

# output data
output <- run_query(getSQL("sql/output.sql"))
output72hrs <- run_query(getSQL("sql/output_72hrs.sql"))

#demographic data separated due to error in processing "numeric" type
demographic <- run_query(getSQL('sql/demographics.sql'))
```

```{r}
intake72hrs <- run_query(getSQL("sql/intake_72hrs.sql"))
output72hrs <- run_query(getSQL("sql/output_72hrs.sql"))
```

# Exclusion criteria 1

Patients with sepsis (not bleeding) 

```{r}
# we want to include only septic patients that are not bleeding.

print('Septic patients:')
nrow(sepsis)

print('Patients bleeding')
nrow(patient_inexcluded_icd9)

patient_included_vector <- sepsis[sepsis$icustay_id %!in% patient_inexcluded_icd9$icustay_id  ,]
patient_included <- data.frame(icustay_id=integer(length(patient_included_vector))) 
patient_included$icustay_id<-patient_included_vector

print('Septic patients that are not bleeding')
nrow(patient_included)

# Exclusion criteria and dataset join (we need new datasets to address exclusion criteria)

selected_cohort<-inner_join(demographic,patient_included)
print('Patients >=16 years old:')
nrow(selected_cohort)

#exclusion based on output amount
## we want to exclude blood output or gastric >= 500 OR stool >= 500 OR emesis >= 500

selected_cohort<-selected_cohort%>%filter(icustay_id %!in% IO_exclusion$ICUSTAY_ID) 

print('Patients with proper amount of output:')
nrow(selected_cohort)

# We are using a left join to join them
vol_leak_index_dataset<-Reduce(function(...) merge(..., all.x=TRUE), list(
   selected_cohort
  ,hct
  ,intake
  ,output
  ,intake72hrs
  ,output72hrs
  ,oasis_related
  ,elixhauser_score
  ,sofa_total_day1_to_day4
))

# Ensures fluid intake is greater than 0.
vol_leak_index_dataset<-sqldf('
SELECT * FROM
vol_leak_index_dataset
WHERE intakes_total > 0 ')

print('patients with fluid intake greater than 0')
nrow(vol_leak_index_dataset)
```

# Variables creation

## Leaking Index

*PLEASE NOTE THE ORDER OF VARIABLES CREATION IS REALLY IMPORTANT SINCE THEY ARE NEEDED FOR EXCLUSION CRITERIA*

```{r}
vol_leak_index_dataset$outputs_total[is.na(vol_leak_index_dataset$outputs_total)] = 0
vol_leak_index_dataset$outputs_total_72[is.na(vol_leak_index_dataset$outputs_total_72)] = 0

# Calculates Fluid Balance
vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
    fluid_balance=(intakes_total-outputs_total),
    fluid_balance_72=(intakes_total_72-outputs_total_72)
  )  

# Ensures fluid balance is greater than 0.
vol_leak_index_dataset<-sqldf('
SELECT * FROM
vol_leak_index_dataset
WHERE fluid_balance > 0 ')

print('patients with fluid balance greater than 0')
nrow(vol_leak_index_dataset)

#Calculates Leak Index
vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
    # old definition (NYU Datathon) leaking_index=((mean_hct_24_36hrs/first_hct_6hrs)-1)*body_surface_area*1561
    leaking_index=((mean_hct_24_36hrs - first_hct_6hrs) / fluid_balance) * body_surface_area
  ) 

# Removes leaking index greater than 0 (means that hematocrit increases over time with fluid). We are not doing this because an increase in hematocrit is clinically possible
# vol_leak_index_dataset<-sqldf('
# SELECT * FROM
# vol_leak_index_dataset
# WHERE leaking_index < 0 ')

print('patients with negative leak index')
nrow(vol_leak_index_dataset)
```

### VLI Distribution

*Without Removing Any Outliers*

```{r}
colname<-names(vol_leak_index_dataset['leaking_index'])  
ifelse( nrow(unique(vol_leak_index_dataset['leaking_index']))<10  
        ,barplot(table(vol_leak_index_dataset['leaking_index']),main=colname,xlab = colname)
        ,hist(as.numeric(unlist(vol_leak_index_dataset['leaking_index']))
        ,main = paste("Histogram of" ,colname),xlab=colname)
         )
cat('Summary of ',colname,'\n',sep = '')  
cat(summary(vol_leak_index_dataset['leaking_index']),'\n',sep = '|')

```

### Addressing VLI outliers

```{r eval=FALSE, include=FALSE}
# We are imputing anything outside the 95% interval
extreme_quants<-as.numeric(quantile(vol_leak_index_dataset$leaking_index, c(.025, .975),na.rm = T))

print('Number of patients with imputed data:')
length(which(vol_leak_index_dataset$leaking_index< extreme_quants[1] | vol_leak_index_dataset$leaking_index> extreme_quants[2]))

medLeak = median(vol_leak_index_dataset$leaking_index, na.rm = T)
vol_leak_index_dataset$leaking_index[which(vol_leak_index_dataset$leaking_index< extreme_quants[1] | vol_leak_index_dataset$leaking_index> extreme_quants[2])] = medLeak
```

*After Removing Any Outliers*

```{r}
colname<-names(vol_leak_index_dataset['leaking_index'])  
ifelse( nrow(unique(vol_leak_index_dataset['leaking_index']))<10  
        ,barplot(table(vol_leak_index_dataset['leaking_index']),main=colname,xlab = colname)
        ,hist(as.numeric(unlist(vol_leak_index_dataset['leaking_index']))
        ,main = paste("Histogram of" ,colname),xlab=colname)
         )
cat('Summary of ',colname,'\n',sep = '')  
cat(summary(vol_leak_index_dataset['leaking_index']),'\n',sep = '|')

```

## SOFA Delta creation

For calculating delta sofa we need tertiles of sofa day 1.

High sofa day 1 = T3
Medium sofa day 1 = T2
Low sofa day 1 = T1


|                   | Sofa day 2 No change | Sofa day 2 Increases | Sofa day 2 Decreases |
|-------------------|----------------------|----------------------|----------------------|
| High Sofa day 1   | Bad                  | Bad                  | Good                 |
| Medium Sofa day 1 | Bad                  | Bad                  | Good                 |
| Low Sofa day 1    | Good                 | Bad                  | Good                 |

0 Means GOOD Oucome, 1 means BAD Outcome

```{r}
vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
     t_sofatotal_day1=as.numeric(cut2(sofatotal_day1, g=3))
    ,delta_sofa=case_when(
      
      t_sofatotal_day1 == 3 & sofatotal_day2 == sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 3 & sofatotal_day2 >  sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 3 & sofatotal_day2 <  sofatotal_day1 ~ 0,
      
      t_sofatotal_day1 == 2 & sofatotal_day2 == sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 2 & sofatotal_day2 >  sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 2 & sofatotal_day2 <  sofatotal_day1 ~ 0,     
      
      t_sofatotal_day1 == 1 & sofatotal_day2 == sofatotal_day1 ~ 0,
      t_sofatotal_day1 == 1 & sofatotal_day2 >  sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 1 & sofatotal_day2 <  sofatotal_day1 ~ 0, 
    )
)
```

# Selecting variables

 - We are just selecting the variables we need for the model or for the exclusion criteria.

```{r}
selected_df <- vol_leak_index_dataset%>%dplyr::select(
     icustay_id
    ,hosp_mortality
    ,hosp_mortality_offset
    ,age_fixed
    ,gender
    ,final_elixhauser_score
    ,oasis
    ,leaking_index
    ,delta_sofa
    ,fluid_balance
    # ,fluid_balance_72
    ,mean_hct_24_36hrs
    ,first_hct_6hrs
  )

selected_df_fluid <- vol_leak_index_dataset%>%dplyr::select(
     icustay_id
    ,hosp_mortality
    ,hosp_mortality_offset
    ,age_fixed
    ,gender
    ,final_elixhauser_score
    ,oasis
    ,leaking_index
    ,delta_sofa
    ,fluid_balance
    ,fluid_balance_72
    ,mean_hct_24_36hrs
    ,first_hct_6hrs
  )

selected_df_fluid<-selected_df_fluid[complete.cases(selected_df_fluid),]
print('Number of patients after removing incomplete data:')
nrow(selected_df_fluid)

# Removing Patients who died before 72 hours
selected_df_fluid<-selected_df_fluid%>%filter(hosp_mortality_offset>72*60 | hosp_mortality_offset==0)
print('Number of patients after excluding patients with hosp_mort_offset <72hrs:')
nrow(selected_df_fluid)

selected_df_fluid<-selected_df_fluid%>%mutate(fluid_balance_72_bin = ifelse(fluid_balance_72 > 0, 1, 0))

# these variables are factors
selected_df$gender<-as.factor(selected_df$gender)

```

# Data Imputation

## Mortality imputation

We are not imputing mortality

# Exclusion Criteria 2


*Please note there are several exclusion criteria and the order in which they are applied is really important.*

 - We are removing patients not in VLI Q2 or VLI Q3.
 - We are removing cases with missing data.

```{r}
print('Number of patients in our dataset so far:')
a<-nrow(selected_df)
a

selected_df<-selected_df%>%filter(!is.na(hosp_mortality))
print('Number of patients excluded due to missing hosp_mortality:')
b<-nrow(selected_df)
a-b

selected_df<-selected_df%>%filter(!is.na(age_fixed))
print('Number of patients excluded due to missing age:')
c<-nrow(selected_df)
b-c

selected_df<-selected_df%>%filter(!is.na(gender))
print('Number of patients excluded due to missing gender:')
d<-nrow(selected_df)
c-d

selected_df<-selected_df%>%filter(!is.na(final_elixhauser_score))
print('Number of patients excluded due to missing final_elixhauser_score:')
e<-nrow(selected_df)
d-e

selected_df<-selected_df%>%filter(!is.na(oasis))
print('Number of patients excluded due to missing oasis:')
f<-nrow(selected_df)
e-f

selected_df<-selected_df%>%filter(!is.na(leaking_index))
print('Number of patients excluded due to missing leaking_index:')
g<-nrow(selected_df)
f-g

selected_df<-selected_df%>%filter(!is.na(delta_sofa))
print('Number of patients excluded due to missing delta_sofa:')
h<-nrow(selected_df)
g-h

('Final number of patients')
h
```

# New Variables Creation

## Leaking Index Quartile

```{r}
selected_df<-selected_df%>%
  mutate(
     q_leaking_index=as.factor(as.numeric(cut2(leaking_index, g=4)))
)
selected_df_fluid<-selected_df_fluid%>%
  mutate(
     q_leaking_index=as.factor(as.numeric(cut2(leaking_index, g=4)))
)
print('cutoff points')
unique(cut2(selected_df$leaking_index, g=4))
```

## Highest_q_leaking_index

```{r}
selected_df<-selected_df%>%mutate(
  highest_q_leaking_index=if_else(q_leaking_index %in% c('4'),1,0 )
)

selected_df$highest_q_leaking_index<-as.factor(selected_df$highest_q_leaking_index)
```

# Exclusion criteria 3

```{r}
#selected_df<-selected_df%>%filter(q_leaking_index %in% c('4','1'))
print('Number of patients after excluding VLI Q2 and VLI Q3:')
nrow(selected_df)
```

# Table 1 

*Stratified by q_leaking_index*

```{r}
mimicTable1 <- table1(~ age_fixed + hosp_mortality + gender + final_elixhauser_score + oasis  + delta_sofa + fluid_balance + mean_hct_24_36hrs + first_hct_6hrs | q_leaking_index, data=selected_df)
mimicTable1
```

# Export dataset

```{r}
write.csv(selected_df,'selected_df.csv')
```

# 72hr Fluid Regression Analysis

```{r}
plot (selected_df$leaking_index, selected_df$fluid_balance_72, xlab = 'leaking_index', ylab = 'fluid balance at 72hrs', main = 'Leaking Index as a predictor of Fluid Balance at 72 hrs')
linearMod <- lm(fluid_balance_72 ~ leaking_index, data=selected_df)
summary(linearMod)
```
