---
title: "01_dataset_creation"
author: "Miguel Ángel Armengol"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_notebook:
    code_folding: hide
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes

knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = paste0(substr(inputFile,1,nchar(inputFile)-4)," ",Sys.Date(),'.html')) })
---

# Environment

```{r message=FALSE, warning=FALSE}
library(bigrquery)
library(summarytools)
library(readr)
library(stringr)
library(sqldf)
library(dplyr)
library(tableone)
library(Hmisc)
library(caret)
library(plotly)
library(table1)
library(Amelia)
library(DescTools)
library(kableExtra)
impute.median <- function(x) replace(x, is.na(x), median(x, na.rm = TRUE))
options(scipen=999)
```


# Set up BigQuery related functions

This chunks also creates the run_query and get_sql function.

```{r setup, include=FALSE}
# Updated for our year
project_id <- "hst-953-2018"
options(httr_oauth_cache=FALSE)
# Function that takes in a sql command and runs it on bigquery
run_query <- function(query){
  data <- query_exec(query, project=project_id, use_legacy_sql=FALSE,max_pages = Inf)
  return(data)
}

# function for reading sql files
getSQL <- function(filepath){
  con = file(filepath, "r")
  sql.string <- ""

  while (TRUE){
    line <- readLines(con, n = 1)

    if ( length(line) == 0 ){
      break
    }

    line <- gsub("\\t", " ", line)

    if(grepl("--",line) == TRUE){
      line <- paste(sub("--","/*",line),"*/")
    }

    sql.string <- paste(sql.string, line)
  }

  close(con)
  return(sql.string)
}

'%!in%' <- function(x,y)!('%in%'(x,y))
```

# Loading queries and extracting the data

Loads all queries from the sql files in the extraction folder and runs them into RBigQuey to extract the data.

```{r}
hct <- run_query(getSQL("sql/hct.sql" ))
# removes missing values
hct<-hct[complete.cases(hct),]

#Remove chronic AKI, creat, rrt patients
# chronicAKI<-run_query(getSQL('sql/aki/chronicAKI.sql'))
# baseline_creat<-run_query(getSQL('sql/aki/baseline_creat.sql'))
# peakcreat48h<-run_query(getSQL('sql/aki/peakcreat48h.sql'))
# peakcreat7days<-run_query(getSQL('sql/aki/peakcreat7days.sql'))

# renal replacement therapy
first_rrt <-run_query(getSQL('sql/aki/first_rrt.sql'))

IO_exclusion <- run_query(getSQL('sql/IO_exclusion.sql'))

charlson_score <- run_query(getSQL('sql/charlson_score.sql'))

#demographic
demographic <- run_query(getSQL('sql/demographics.sql'))

# apache related variables
apache_related <- run_query(getSQL('sql/apache_related.sql'))

# IMPORTANT, ACCORDING TO THE DOCUMMENTATION -1 in Apache score means missing.
apache_related$predictedHospitalMortality[apache_related$predictedHospitalMortality==(-1)] <- NA
apache_related$apache_iv[apache_related$apache_iv==(-1)] <- NA
apache_related$apachescore[apache_related$apachescore==(-1)] <- NA


# SOFA related variables
sofa_cv_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_cv_day1_to_day4.sql"))
sofa_renal_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_renal_day1_to_day4.sql"))
sofa_respi_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_respi_day1_to_day4.sql"))
sofa_3others_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_3others_day1_to_day4.sql"))
sofa_total_day1_to_day4 <- sqldf((getSQL("sql/sofa/sofa_total_day1_to_day4.sql")))

# List of patients that are bleeding we want to exclude.
patient_inexcluded_icd9 <- run_query(getSQL('sql/patient_inexcluded_icd9.sql'))

# Sepsis patients according to angus criteria
sepsis <- run_query(getSQL('sql/sepsis.sql'))

# Fluid data
intake <- run_query(getSQL("sql/intake.sql"))

# 72hr fluid data
intake72hrs <- run_query(getSQL("sql/intake72hrs.sql"))

# mech ventilated patients
vent <- run_query((getSQL('sql/vent.sql')))

#temperature
temp <- run_query((getSQL('sql/temp.sql')))
```

```{r}
# Separate fluid intake calls ("full" fluid means all fluid for a patient, not just
# crystalloids). We are not using the full fluid intakes. We are only interested in crystalloiods.
intake <- run_query(getSQL("sql/intake.sql"))
intake_full <-run_query(getSQL("sql/intake_full.sql"))
first_rrt <-run_query(getSQL('sql/aki/first_rrt.sql'))
```


# Exclusion criteria 1

Patients with sepsis (not bleeding) 

```{r}
# we want to include septic patients that are not bleeding.

print('Septic patients:')
nrow(sepsis)

print('Patients bleeding')
nrow(patient_inexcluded_icd9)

patient_included_vector <- sepsis[sepsis$patientunitstayid %!in% patient_inexcluded_icd9$patientUnitStayID  ,]
patient_included <- data.frame(patientunitstayid=integer(length(patient_included_vector))) 
patient_included$patientunitstayid<-patient_included_vector

print('Septic patients that are not bleeding')
nrow(patient_included)

selected_cohort<-inner_join(demographic,patient_included)
print('Patients >=16 years old:')
nrow(selected_cohort)

#exlusion based on output amount
## we want to exclude gastric >= 500 OR stool >= 500 OR emesis >= 500

#Exclusion criteria and dataset join (we need new datasets to address exclusion criteria)

selected_cohort<-sqldf('
SELECT * FROM 
selected_cohort
LEFT JOIN
IO_exclusion
USING
(patientunitstayid)
WHERE
IO_exclusion.patientunitstayid IS NULL
')

selected_cohort<-sqldf('
SELECT * FROM 
selected_cohort
LEFT JOIN
first_rrt
USING
(patientunitstayid)
WHERE
first_rrt.patientunitstayid IS NULL
')

print('Patients without rrt or expulsion of fluids:')
nrow(selected_cohort)

# We are using a left join to join them
vol_leak_index_dataset<-Reduce(function(...) merge(..., all.x=TRUE), list(
   selected_cohort
  ,hct
  ,intake
  ,intake72hrs
  ,apache_related
  ,charlson_score
  ,sofa_total_day1_to_day4
  ,vent
))

# Removes Patients without intake data
vol_leak_index_dataset<-sqldf('
SELECT * FROM
vol_leak_index_dataset
WHERE intakes_total IS NOT null')

print('patients with reliable fluid data')
nrow(vol_leak_index_dataset)
```

# Variables creation

## Leaking Index

*PLEASE NOTE THE ORDER OF VARIABLES CREATION IS REALLY IMPORTANT SINCE THEY ARE NEEDED FOR EXCLUSION CRITERIA*

```{r}
vol_leak_index_dataset$outputs_total[is.na(vol_leak_index_dataset$outputs_total)] = 0
vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
    fluid_balance=(intakes_total-outputs_total),
    fluid_balance_72 = (intakes_total72 - outputs_total72)
  )  
hist(vol_leak_index_dataset['fluid_balance'])
hist(vol_leak_index_dataset['fluid_balance_72'])

# Ensures fluid balance is greater than 0.
vol_leak_index_dataset<-sqldf('
SELECT * FROM
vol_leak_index_dataset
WHERE fluid_balance > 0 ')

print('patients with fluid balance greater than 0')
nrow(vol_leak_index_dataset)

vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
    # leaking_index=((mean_hct_24_36hrs/first_hct_6hrs)-1)*body_surface_area*1561
    leaking_index=((mean_hct_24_36hrs - first_hct_6hrs) / fluid_balance) * body_surface_area
  )  

# Removes leaking index greater than 0 (means that hematocrit increases over time with fluid). We are not doing this now.
# vol_leak_index_dataset<-sqldf('
# SELECT * FROM
# vol_leak_index_dataset
# WHERE leaking_index < 0 ')


```


### VLI Distribution

*Before addressing outliers*

```{r}
colname<-names(vol_leak_index_dataset['leaking_index'])  
ifelse( nrow(unique(vol_leak_index_dataset['leaking_index']))<10  
        ,barplot(table(vol_leak_index_dataset['leaking_index']),main=colname,xlab = colname)
        ,hist(as.numeric(unlist(vol_leak_index_dataset['leaking_index']))
        ,main = paste("Histogram of" ,colname),xlab=colname)
         )
cat('Summary of ',colname,'\n',sep = '')  
cat(summary(vol_leak_index_dataset['leaking_index']),'\n',sep = '|')

```

### Addressing VLI and Fluid outliers
Imputing values outside 95% interval with median

```{r eval=FALSE, include=FALSE}

extreme_quants<-as.numeric(quantile(vol_leak_index_dataset$leaking_index, c(.025, .975),na.rm = T))

print('Number of patients with imputed data:')
length(which(vol_leak_index_dataset$leaking_index< extreme_quants[1] | vol_leak_index_dataset$leaking_index> extreme_quants[2]))

medLeak = median(vol_leak_index_dataset$leaking_index, na.rm = T)
vol_leak_index_dataset$leaking_index[which(vol_leak_index_dataset$leaking_index< extreme_quants[1] | vol_leak_index_dataset$leaking_index> extreme_quants[2])] = medLeak
```

### VLI Distribution after removing outliers

*After addressing outliers*

```{r}
colname<-names(vol_leak_index_dataset['leaking_index'])  
ifelse( nrow(unique(vol_leak_index_dataset['leaking_index']))<10  
        ,barplot(table(vol_leak_index_dataset['leaking_index']),main=colname,xlab = colname)
        ,hist(as.numeric(unlist(vol_leak_index_dataset['leaking_index']))
        ,main = paste("Histogram of" ,colname),xlab=colname)
         )
cat('Summary of ',colname,'\n',sep = '')  
cat(summary(vol_leak_index_dataset['leaking_index']),'\n',sep = '|')


```

## SOFA Delta and others creation

For calculating delta sofa we need tertiles of sofa day 1.

High sofa day 1 = T3
Medium sofa day 1 = T2
Low sofa day 1 = T1


|                   | Sofa day 2 No change | Sofa day 2 Increases | Sofa day 2 Decreases |
|-------------------|----------------------|----------------------|----------------------|
| High Sofa day 1   | Bad                  | Bad                  | Good                 |
| Medium Sofa day 1 | Bad                  | Bad                  | Good                 |
| Low Sofa day 1    | Good                 | Bad                  | Good                 |

0 Means GOOD Oucome, 1 means BAD Outcome

```{r}
vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
     t_sofatotal_day1=as.numeric(cut2(sofatotal_day1, g=3))
    ,delta_sofa=case_when(
      
      t_sofatotal_day1 == 3 & sofatotal_day2 == sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 3 & sofatotal_day2 >  sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 3 & sofatotal_day2 <  sofatotal_day1 ~ 0,
      
      t_sofatotal_day1 == 2 & sofatotal_day2 == sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 2 & sofatotal_day2 >  sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 2 & sofatotal_day2 <  sofatotal_day1 ~ 0,     
      
      t_sofatotal_day1 == 1 & sofatotal_day2 == sofatotal_day1 ~ 0,
      t_sofatotal_day1 == 1 & sofatotal_day2 >  sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 1 & sofatotal_day2 <  sofatotal_day1 ~ 0, 
    )
    ,sofatotal_day1_cat=case_when(
      
      between(sofatotal_day1,0,7) ~ 'a) 0-7',
      between(sofatotal_day1,8,11) ~ 'b) 8-11',
      sofatotal_day1>=12 ~ 'c) >=12'
    )    
)
```

# Selecting variables

 - We are just selecting the variables we need for the model or for the exclusion criteria.

```{r}
selected_df <- vol_leak_index_dataset%>%dplyr::select(
     patientunitstayid
    ,hosp_mortality
    ,hosp_mortality_offset
    ,age_fixed
    ,gender
    ,final_charlson_score
    ,apache_iv
    ,leaking_index
    ,sofatotal_day1
    ,sofatotal_day1_cat
    ,delta_sofa
    ,fluid_balance
    # ,fluid_balance_72
    ,mean_hct_24_36hrs
    ,first_hct_6hrs
    ,mech_vent
    , actualicumortality
    , unabridgedunitlos
    , unabridgedhosplos    
  )

selected_df_fluid <- vol_leak_index_dataset%>%dplyr::select(
     patientunitstayid
    ,hosp_mortality
    ,hosp_mortality_offset
    ,age_fixed
    ,gender
    ,final_charlson_score
    ,apache_iv
    ,leaking_index
    ,delta_sofa
    ,fluid_balance
    ,fluid_balance_72
    ,mean_hct_24_36hrs
    ,first_hct_6hrs
  )
# these variables are factors
```

# Data Imputation

## Mortality imputation

*TODO: Improve hospital mortality extraction, don't impute*

# Exclusion Criteria 2


*Please note there are several exclusion criteria and the order in which they are applied is really important.*

 - We are removing patients not in VLI Q2 or VLI Q3.
 - We are removing cases with missing data.



```{r}
print('Number of patients in our dataset so far:')
nrow(selected_df)

selected_df<-selected_df[complete.cases(selected_df),]
print('Number of patients after excluding patients with missing data:')
nrow(selected_df)

#  we don’t want to remove patients that died before 72hrs for the mortality/delta_sofa analysis. This removes a significant number of patients who passed away from our analysis and we don’t see as strong of a signal.


selected_df_fluid<-selected_df_fluid[complete.cases(selected_df_fluid),]
print('Number of patients after excluding patients with missing data:')
nrow(selected_df_fluid)

selected_df_fluid<-selected_df_fluid%>%filter(hosp_mortality_offset>72*60 | hosp_mortality_offset==0)
print('Number of patients after excluding patients with hosp_mort_offset <72hrs:')
nrow(selected_df_fluid)
```

# New Variables Creation

## Leaking Index Quartile

```{r}
selected_df<-selected_df%>%
  mutate(
     q_leaking_index=as.factor(as.numeric(cut2(leaking_index, g=4)))
)

selected_df_fluid<-selected_df_fluid%>%
  mutate(
     q_leaking_index=as.factor(as.numeric(cut2(leaking_index, g=4)))
)
```

## Highest_q_leaking_index

```{r}
selected_df<-selected_df%>%mutate(
  highest_q_leaking_index=if_else(q_leaking_index %in% c('4'),1,0 )
)

selected_df$highest_q_leaking_index<-as.factor(selected_df$highest_q_leaking_index)

print('cutoff points')
unique(cut2(selected_df$leaking_index, g=4))
```

# Exclusion criteria 3

Not used
```{r eval=FALSE, include=FALSE}
# selected_df<-selected_df%>%filter(q_leaking_index %in% c('4','1'))
print('Number of patients after excluding VLI Q2 and VLI Q3:')
nrow(selected_df)
```

# Table 1 

```{r}
vars_in_table1<-c('age_fixed', 'gender'  , 'final_charlson_score' , 'apache_iv'  , 'delta_sofa' ,'sofatotal_day1_cat', 'fluid_balance' , 'mean_hct_24_36hrs' , 'first_hct_6hrs', 'mech_vent','hosp_mortality_offset', 'actualicumortality' , 'unabridgedunitlos' , 'unabridgedhosplos',
                  'leaking_index')

table1_dataset<-selected_df[,vars_in_table1]

cat_variables<-rep(NA, length(vars_in_table1))
stratifyby<-"sofatotal_day1_cat"


label(table1_dataset$age_fixed)<-'Age'

# detects whether a variable is categorical or not
cont<-1
for (i in 1:length(vars_in_table1) ) {
  if ( n_distinct(table1_dataset[vars_in_table1[i] ])<=10 ) {
    print(i)
    print(vars_in_table1[i])
    print(names(table1_dataset[vars_in_table1[i]]))
    cat_variables[cont]<-names(table1_dataset[vars_in_table1[i]])
    cont<-cont+1
  }
}  

cat_variables<-cat_variables[!is.na(cat_variables)]

table1_base<-print(CreateTableOne(vars = vars_in_table1, strata = stratifyby, factorVars = cat_variables
    ,data = table1_dataset, addOverall=T),varLabels = T)

# run this in console for html output, the code below uses kableExtra::

starification_cats<-n_distinct(table1_dataset[,stratifyby])

table1_base %>%
  kbl(caption = "Table 1 of base model" , align = "c") %>%
  kable_classic_2(full_width = F, html_font = "Cambria")%>%
  add_header_above(c(" "," ", 'SOFA day1 Category' = starification_cats," ", "" ))
```



# Export dataset

```{r}
write.csv(selected_df,'selected_df.csv')
```

# Fluid Balance Regression Analysis

```{r}
plot (vol_leak_index_dataset$fluid_balance, vol_leak_index_dataset$hosp_mortality, xlim = c(0, 25000))
linearMod <- lm(hosp_mortality ~ fluid_balance, data=vol_leak_index_dataset)
summary(linearMod)
```
# 72hr Fluid Regression Analysis
```{r}
plot (vol_leak_index_dataset$leaking_index, vol_leak_index_dataset$fluid_balance_72, xlab = 'leaking_index', ylab = 'fluid balance at 72hrs', main = 'Leaking Index as a predictor of Fluid Balance at 72 hrs')
linearMod <- lm(fluid_balance_72 ~ leaking_index, data=vol_leak_index_dataset)
summary(linearMod)
```

# 72hr Fluid Regression Based on Quartiles
```{r}
plot (selected_df$q_leaking_index, selected_df$fluid_balance_72, xlab = 'leaking_index', ylab = 'fluid balance at 72hrs', main = 'Leaking Index as a predictor of Fluid Balance at 72 hrs')
linearMod <- lm(fluid_balance_72 ~ q_leaking_index + age_fixed + gender + fluid_balance + apache_iv + final_charlson_score, data=selected_df)
summary(linearMod)
```
