---
title: "01_dataset_creation"
author: "Jay Chandra & Miguel Ángel Armengol"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_notebook:
    code_folding: hide
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes

knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = paste0(substr(inputFile,1,nchar(inputFile)-4)," ",Sys.Date(),'.html')) })
---

# Environment

```{r message=FALSE, warning=FALSE}
library(bigrquery)
library(summarytools)
library(readr)
library(stringr)
library(sqldf)
library(dplyr)
library(tableone)
library(Hmisc)
library(kableExtra)
library(caret)
library(plotly)
library(table1)

winsorize_x = function(x, cut = 0.05){
  cut_point_top <- quantile(x, 1 - cut, na.rm = T)
  cut_point_bottom <- quantile(x, cut, na.rm = T)
  i = which(x >= cut_point_top) 
  x[i] = cut_point_top
  j = which(x <= cut_point_bottom) 
  x[j] = cut_point_bottom
  return(x)
}
```


# Set up BigQuery related functions

This chunks also creates the run_query and get_sql function.

```{r setup, include=FALSE}
# Updated for our year
project_id <- "hst-953-2019"
options(httr_oauth_cache=FALSE)
# Function that takes in a sql command and runs it on bigquery
run_query <- function(query){
  data <- query_exec(query, project=project_id, use_legacy_sql=FALSE,max_pages = Inf)
  return(data)
}

# function for reading sql files
getSQL <- function(filepath){
  con = file(filepath, "r")
  sql.string <- ""

  while (TRUE){
    line <- readLines(con, n = 1)

    if ( length(line) == 0 ){
      break
    }

    line <- gsub("\\t", " ", line)

    if(grepl("--",line) == TRUE){
      line <- paste(sub("--","/*",line),"*/")
    }

    sql.string <- paste(sql.string, line)
  }

  close(con)
  return(sql.string)
}

'%!in%' <- function(x,y)!('%in%'(x,y))
```

# Loading queries and extracting the data

Loads all queries from the sql files in the extraction folder and runs them into RBigQuery to extract the data.

```{r}
# overall characteristics
amsterdam = run_query(getSQL("sql/amsterdam_hct.sql" ))

# Sepsis patients
sepsis <- run_query(getSQL('sql/sepsis_new.sql'))

# List of patients that are bleeding we want to exclude.
patient_inexcluded_icd9 <- run_query(getSQL('sql/patient_inexcluded_icd9.sql'))

#exclude patients who have recieved blood
blood_infus <- run_query(getSQL('sql/blood_infus.sql'))

#72hr Fluid Data
fluid_72hrs <- run_query(getSQL('sql/fluid_72hrs.sql'))

dynamicFluid <- run_query(getSQL('sql/dynamicFluid.sql'))
dynamicFluid_72 <- run_query(getSQL('sql/dynamicFluid_72.sql'))

#Renal Replacement Therapy
rrt <- run_query(getSQL('sql/RRT.sql'))

apacheii <- read.csv(file = 'apache_ii.csv')
apacheii <- apacheii %>% select(admissionid, apache_ii_total_score)
apacheii <- apacheii %>% rename(apache = apache_ii_total_score)
```

# Exclusion criteria 1

Patients with sepsis (not bleeding)

```{r}
# we want to include septic patients that are not bleeding or recieving blood

print('Septic patients:')
nrow(sepsis)

patient_included_vector <- sepsis[sepsis$admissionid %!in% patient_inexcluded_icd9$admissionid & sepsis$admissionid %!in% blood_infus$admissionid ,]
patient_included <- data.frame(patientunitstayid=integer(length(patient_included_vector))) 
patient_included$patientunitstayid<-patient_included_vector

print('Septic patients that are not bleeding or recieving blood')
nrow(patient_included)

```


# Exclusion criteria 2 and JOIN

Exclusion criteria and dataset join and not getting rrt (we need new datasets to address exclusion criteria)

```{r}
names(amsterdam)[1] <- "admissionid"
amsterdam$unitType <- "mixed surgical-medical"
names(patient_included)[1] <- "admissionid"
selected_cohort<-inner_join(amsterdam,patient_included)

selected_cohort<-sqldf('
SELECT * FROM
selected_cohort
LEFT JOIN
rrt
USING
(admissionid)
WHERE
rrt.admissionid IS NULL
')

print('Patients not recieving rrt:')
nrow(selected_cohort)

# We are using a left join to join them
vol_leak_index_dataset<-Reduce(function(...) merge(..., all.x=TRUE), list(
   selected_cohort
   ,dynamicFluid
   ,dynamicFluid_72
   ,fluid_72hrs
  ,charlson
  ,apacheii
))

vol_leak_index_dataset$intakes_72[is.na(vol_leak_index_dataset$intakes_72)] = 0
vol_leak_index_dataset$fluidin_72[is.na(vol_leak_index_dataset$fluidin_72)] = 0
vol_leak_index_dataset$intakes[is.na(vol_leak_index_dataset$intakes)] = 0
vol_leak_index_dataset$fluidin[is.na(vol_leak_index_dataset$fluidin)] = 0

# We are adding mech vent status
vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
    mech_vent = ifelse(unabridgedactualventdays > 0, 1, 0),
    intakes = fluidin + intakes,
    intakes_72 = fluidin_72 + intakes_72,
    totalFluid = intakes - fluidout,
    totalFluid_72 = intakes_72 - fluidout_72
  )
vol_leak_index_dataset$mech_vent[is.na(vol_leak_index_dataset$mech_vent)] = 0
# Removes Patients without intake data
vol_leak_index_dataset<-sqldf('
SELECT * FROM
vol_leak_index_dataset
WHERE intakes IS NOT null')
print('Not missing fluid data')
nrow(vol_leak_index_dataset)

# Ensures fluid balance is positive.
vol_leak_index_dataset<-sqldf('
SELECT * FROM
vol_leak_index_dataset
WHERE totalFluid > 0 ')

print('patients with positive fluid data')
nrow(vol_leak_index_dataset)
```

## Creating new variables

### Leaking Index

```{r}
# 100 is for a unit conversion of hematocrit
vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
    leaking_index=((mean_hct_24_36hrs - first_hct_6hrs) * (100) / totalFluid) * body_surface_area * 1000
  )  
```

### Addressing outliers

```{r eval=FALSE, include=FALSE}

# We are removing VLI outliers
# We are imputing anything outside the 90% interval
extreme_quants<-as.numeric(quantile(vol_leak_index_dataset$leaking_index, c(.05, .95),na.rm = T))

print('Number of patients with imputed data:')
length(which(vol_leak_index_dataset$leaking_index< extreme_quants[1] | vol_leak_index_dataset$leaking_index> extreme_quants[2]))

medLeak = median(vol_leak_index_dataset$leaking_index, na.rm = T)
vol_leak_index_dataset$leaking_index[which(vol_leak_index_dataset$leaking_index< extreme_quants[1] | vol_leak_index_dataset$leaking_index> extreme_quants[2])] = medLeak
```

## SOFA Delta Description

|                   | Sofa day 2 No change | Sofa day 2 Increases | Sofa day 2 Decreases |
|-------------------|----------------------|----------------------|----------------------|
| High Sofa day 1   | Bad                  | Bad                  | Good                 |
| Medium Sofa day 1 | Bad                  | Bad                  | Good                 |
| Low Sofa day 1    | Good                 | Bad                  | Good                 |

0 Means GOOD Oucome, 1 means BAD Outcome

# Creating and Altering Variable
```{r}
vol_leak_index_dataset$hosp_mortality = vol_leak_index_dataset$actualhospitalmortality * 1
vol_leak_index_dataset$q_leaking_index = as.factor(ntile(vol_leak_index_dataset$leaking_index,4))
```

## Selecting/Imputing/removing data

```{r}
selected_df_fluid <- vol_leak_index_dataset%>%dplyr::select(
     admissionid
    ,hosp_mortality
    ,age_fixed
    ,gender
    ,hosp_mortality_offset
    ,totalFluid
    ,totalFluid_72
    ,q_leaking_index
    ,leaking_index
    ,first_hct_6hrs
    ,mean_hct_24_36hrs
    , unabridgedunitlos
    , unabridgedhosplos
    , unitType
    ,apache
    ,weight
    ,height
    ,body_surface_area
  )

selected_df_mortality <- vol_leak_index_dataset%>%dplyr::select(
     admissionid
    ,hosp_mortality
    ,age_fixed
    ,gender
    ,hosp_mortality_offset
    ,totalFluid
    ,q_leaking_index
    ,leaking_index
    ,first_hct_6hrs
    ,mean_hct_24_36hrs
    , unabridgedunitlos
    , unabridgedhosplos
    , unitType
    ,apache
    ,weight
    ,height
    ,body_surface_area
  )

selected_df_fluid = selected_df_fluid[selected_df_fluid$gender != "", ]
selected_df_mortality = selected_df_mortality[selected_df_mortality$gender != "", ]

selected_df_fluid<-selected_df_fluid[complete.cases(selected_df_fluid),]
print('Number of patients after removing incomplete data (fluid):')
nrow(selected_df_fluid)

# Removing Patients who died before 84 hours
selected_df_fluid<-selected_df_fluid%>%filter(hosp_mortality_offset>84 | hosp_mortality_offset==0)
print('Number of patients after excluding patients with hosp_mort_offset <84hrs:')
nrow(selected_df_fluid)

selected_df_mortality<-selected_df_mortality[complete.cases(selected_df_mortality),]
print('Number of patients after removing incomplete data (mortality):')
nrow(selected_df_mortality)

table(cut2(selected_df_fluid$leaking_index, g=4))
table(cut2(selected_df_mortality$leaking_index, g=4))
```

# Table 1

```{r}
table1(~ gender + age_fixed + hosp_mortality + totalFluid + mean_hct_24_36hrs + first_hct_6hrs| q_leaking_index, data=selected_df_mortality)
```
## Summary of dataset
```{r}
view(dfSummary(selected_df_fluid))
```

## Final Histogram of Fluid Balance
```{r}
hist(selected_df_fluid$totalFluid, breaks = 50, xlim = c(0, 10000))
```

```{r}
hist(selected_df_fluid$totalFluid_72, breaks = 50)
```

```{r}
png("amsterdam_hist.png", width = 2000, height = 2000, res = 400)
hist(selected_df_mortality$leaking_index, main = "Amsterdam",  xlab = "VLI", ylab = "Frequency", xlim = c(-15, 5), breaks = 15)
dev.off()
```


```{r}
write.csv(selected_df_fluid,'selected_df_fluid.csv')
write.csv(selected_df_mortality,'selected_df_mortality.csv')
```


