---
title: "01_dataset_creation"
author: "Miguel Ángel Armengol"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_notebook:
    code_folding: hide
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes

knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = paste0(substr(inputFile,1,nchar(inputFile)-4)," ",Sys.Date(),'.html')) })
---

# Environment

```{r message=FALSE, warning=FALSE}
library(bigrquery)
library(summarytools)
library(readr)
library(stringr)
library(sqldf)
library(dplyr)
library(tableone)
library(Hmisc)
library(caret)
library(plotly)
library(table1)
impute.median <- function(x) replace(x, is.na(x), median(x, na.rm = TRUE))
```


# Set up BigQuery related functions

This chunks also creates the run_query and get_sql function.

```{r setup, include=FALSE}
# Updated for our year
project_id <- "hst-953-2018"
options(httr_oauth_cache=FALSE)
# Function that takes in a sql command and runs it on bigquery
run_query <- function(query){
  data <- query_exec(query, project=project_id, use_legacy_sql=FALSE,max_pages = Inf)
  return(data)
}

# function for reading sql files
getSQL <- function(filepath){
  con = file(filepath, "r")
  sql.string <- ""

  while (TRUE){
    line <- readLines(con, n = 1)

    if ( length(line) == 0 ){
      break
    }

    line <- gsub("\\t", " ", line)

    if(grepl("--",line) == TRUE){
      line <- paste(sub("--","/*",line),"*/")
    }

    sql.string <- paste(sql.string, line)
  }

  close(con)
  return(sql.string)
}

'%!in%' <- function(x,y)!('%in%'(x,y))
```

# Loading queries and extracting the data

Loads all queries from the sql files in the extraction folder and runs them into RBigQuey to extract the data.

```{r}
hct <- run_query(getSQL("sql/hct.sql" ))
# removes missing values
hct<-hct[complete.cases(hct),]

#Remove chronic AKI, creat, rrt patients
# chronicAKI<-run_query(getSQL('sql/aki/chronicAKI.sql'))
# baseline_creat<-run_query(getSQL('sql/aki/baseline_creat.sql'))
# peakcreat48h<-run_query(getSQL('sql/aki/peakcreat48h.sql'))
# peakcreat7days<-run_query(getSQL('sql/aki/peakcreat7days.sql'))

# renal replacement therapy
# first_rrt <-run_query(getSQL('sql/aki/first_rrt.sql'))

IO_exclusion <- run_query(getSQL('sql/IO_exclusion.sql'))

charlson_score <- run_query(getSQL('sql/charlson_score.sql'))

# gathers patientunitstayids with reliable fluid data
all_reliable <- run_query (getSQL("sql/all_reliable.sql"))

#demographic
demographic <- run_query(getSQL('sql/demographics.sql'))

# apache related variables
apache_related <- run_query(getSQL('sql/apache_related.sql'))

# IMPORTANT, ACCORDING TO THE DOCUMMENTATION -1 in Apache score means missing.
apache_related$predictedHospitalMortality[apache_related$predictedHospitalMortality==(-1)] <- NA
apache_related$apache_iv[apache_related$apache_iv==(-1)] <- NA
apache_related$apachescore[apache_related$apachescore==(-1)] <- NA


# SOFA related variables
sofa_cv_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_cv_day1_to_day4.sql"))
sofa_renal_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_renal_day1_to_day4.sql"))
sofa_respi_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_respi_day1_to_day4.sql"))
sofa_3others_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_3others_day1_to_day4.sql"))

# List of patients that are bleeding we want to exclude.
patient_inexcluded_icd9 <- run_query(getSQL('sql/patient_inexcluded_icd9.sql'))

# Sepsis patients according to angus criteria
sepsis <- run_query(getSQL('sql/sepsis.sql'))

# Fluid data
fluidIntake <- run_query(getSQL("sql/intake_output.sql"))
```

## Fluid intake calculation

```{r}
fluidIntake<-fluidIntake%>%
  mutate( totalFluid = (intakes-outputs))
```

# Exclusion criteria 1

Patients with sepsis (not bleeding) 

```{r}
# we want to include septic patients that are not bleeding.

print('Septic patients:')
nrow(sepsis)

print('Patients bleeding')
nrow(patient_inexcluded_icd9)

patient_included_vector <- sepsis[sepsis$patientunitstayid %!in% patient_inexcluded_icd9$patientUnitStayID  ,]
patient_included <- data.frame(patientunitstayid=integer(length(patient_included_vector))) 
patient_included$patientunitstayid<-patient_included_vector

print('Septic patients that are not bleeding')
nrow(patient_included)

```

# Exclusion criteria 2

Exclusion criteria and dataset join (we need new datasets to address exclusion criteria)

```{r}
selected_cohort<-inner_join(demographic,patient_included)
print('Patients >=16 years old:')
nrow(selected_cohort)

#exlusion based on output amount
## we want to exclude gastric >= 500 OR stool >= 500 OR emesis >= 500

selected_cohort<-sqldf('
SELECT * FROM 
selected_cohort
LEFT JOIN
IO_exclusion
USING
(patientunitstayid)
WHERE
IO_exclusion.patientunitstayid IS NULL
')

print('Patients with proper amount of output:')
nrow(selected_cohort)

# We are using a left join to join them
vol_leak_index_dataset<-Reduce(function(...) merge(..., all.x=TRUE), list(
   selected_cohort
  ,hct
  ,fluidIntake
  ,apache_related
  ,charlson_score
  ,sofa_total_day1_to_day4
))

# Ensures that patients have reliable fluid data
# we have a list with reliable fluid data and we will use only patients in this list
vol_leak_index_dataset<-sqldf('
SELECT * FROM
vol_leak_index_dataset
INNER JOIN
all_reliable
USING(patientunitstayid)
')
print('patients with reliable fluid data')
nrow(vol_leak_index_dataset)

# vol_leak_index_dataset<-sqldf('
# SELECT * FROM
# vol_leak_index_dataset
# INNER JOIN
# all_reliable
# USING(patientunitstayid)
# ')

# Ensures fluid intake is greater than 0.
vol_leak_index_dataset<-sqldf('
SELECT * FROM
vol_leak_index_dataset
WHERE totalFluid > 0 ')

print('patients with fluid intake greater than 0')
nrow(vol_leak_index_dataset)
```

# Variables creation

## Leaking Index

*PLEASE NOTE THE ORDER OF VARIABLES CREATION IS REALLY IMPORTANT SINCE THEY ARE NEEDED FOR EXCLUSION CRITERIA*

```{r}
vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
    # leaking_index=((mean_hct_24_36hrs/first_hct_6hrs)-1)*body_surface_area*1561
    leaking_index=((mean_hct_24_36hrs - first_hct_6hrs) / totalFluid) * body_surface_area
  )  
```

### VLI Distribution

```{r}
colname<-names(vol_leak_index_dataset['leaking_index'])  
ifelse( nrow(unique(vol_leak_index_dataset['leaking_index']))<10  
        ,barplot(table(vol_leak_index_dataset['leaking_index']),main=colname,xlab = colname)
        ,hist(as.numeric(unlist(vol_leak_index_dataset['leaking_index']))
        ,main = paste("Histogram of" ,colname),xlab=colname)
         )
cat('Summary of ',colname,'\n',sep = '')  
cat(summary(vol_leak_index_dataset['leaking_index']),'\n',sep = '|')

```

### Addressing outliers

```{r}
## Median imputation for outliers
outliers <- boxplot(vol_leak_index_dataset$leaking_index, plot=FALSE)$out 
vol_leak_index_dataset$leaking_index[which(vol_leak_index_dataset$leaking_index %in% outliers)]<-median(vol_leak_index_dataset$leaking_index)

## Windsorization Clean data by means of winsorization, i.e., by shrinking outlying observations to the border of the main part of the data.
#vol_leak_index_dataset<-vol_leak_index_dataset%>%mutate(leaking_index=winsorize(leaking_index))
```

### Data Imputation

```{r}
vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
    # leaking_index=((mean_hct_24_36hrs/first_hct_6hrs)-1)*body_surface_area*1561
    leaking_index=impute.median(leaking_index)
  )  
```

## Derived Leaking Index Variables

```{r}
vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
     q_leaking_index=as.factor(as.numeric(cut2(leaking_index, g=4)))
    ,bin_leaking_index=if_else(q_leaking_index==4,1,0)
)
```

## SOFA Delta creation

For calculating delta sofa we need tertiles of sofa day 1.

High sofa day 1 = T3
Medium sofa day 1 = T2
Low sofa day 1 = T1


|                   | Sofa day 2 No change | Sofa day 2 Increases | Sofa day 2 Decreases |
|-------------------|----------------------|----------------------|----------------------|
| High Sofa day 1   | Bad                  | Bad                  | Good                 |
| Medium Sofa day 1 | Bad                  | Bad                  | Good                 |
| Low Sofa day 1    | Good                 | Bad                  | Good                 |

0 Means GOOD Oucome, 1 means BAD Outcome

```{r}
vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
     t_sofatotal_day1=as.numeric(cut2(sofatotal_day1, g=3))
    ,delta_sofa=case_when(
      
      t_sofatotal_day1 == 3 & sofatotal_day2 == sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 3 & sofatotal_day2 >  sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 3 & sofatotal_day2 <  sofatotal_day1 ~ 0,
      
      t_sofatotal_day1 == 2 & sofatotal_day2 == sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 2 & sofatotal_day2 >  sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 2 & sofatotal_day2 <  sofatotal_day1 ~ 0,     
      
      t_sofatotal_day1 == 1 & sofatotal_day2 == sofatotal_day1 ~ 0,
      t_sofatotal_day1 == 1 & sofatotal_day2 >  sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 1 & sofatotal_day2 <  sofatotal_day1 ~ 0, 
    )
)
```

# Selecting variables

 - We are just selecting the variables we need for the model or for the exclusion criteria.

```{r}
selected_df <- vol_leak_index_dataset%>%select(
     patientunitstayid
    ,hosp_mortality
    ,age_fixed
    ,gender
    ,final_charlson_score
    ,apache_iv
    ,leaking_index
    ,q_leaking_index
    ,delta_sofa
  )
# these variables are factors
```

# Exclusion Criteria 3

*TODO: Improve hospital mortality, impute apache *

*Please note there are several exclusion criteria and the order were they are applied is really important.*

 - We are removing patients not in VLI Q2 or VLI Q3.
 - We are removing cases with missing data.

```{r}
print('Number of patients in our dataset so far:')
nrow(selected_df)

selected_df<-selected_df[complete.cases(selected_df),]
print('Number of patients after excluding patients with missing data:')
nrow(selected_df)

selected_df<-selected_df%>%filter(q_leaking_index %in% c('4','1'))
print('Number of patients after excluding VLI Q2 and VLI Q3:')
nrow(selected_df)
```

# Export dataset

```{r}
write.csv(selected_df,'selected_df.csv')
```
