---
title: "01_dataset_creation"
author: "Miguel Ángel Armengol"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_notebook:
    code_folding: hide
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes

knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = paste0(substr(inputFile,1,nchar(inputFile)-4)," ",Sys.Date(),'.html')) })
---

# Environment

```{r message=FALSE, warning=FALSE}
library(bigrquery)
library(summarytools)
library(readr)
library(stringr)
library(sqldf)
library(dplyr)
library(tableone)
library(Hmisc)
library(caret)
library(plotly)
```


# Set up BigQuery related functions

This chunks also creates the run_query and get_sql function.

```{r setup, include=FALSE}
# Updated for our year
project_id <- "hst-953-2018"
options(httr_oauth_cache=FALSE)
# Function that takes in a sql command and runs it on bigquery
run_query <- function(query){
  data <- query_exec(query, project=project_id, use_legacy_sql=FALSE,max_pages = Inf)
  return(data)
}

# function for reading sql files
getSQL <- function(filepath){
  con = file(filepath, "r")
  sql.string <- ""

  while (TRUE){
    line <- readLines(con, n = 1)

    if ( length(line) == 0 ){
      break
    }

    line <- gsub("\\t", " ", line)

    if(grepl("--",line) == TRUE){
      line <- paste(sub("--","/*",line),"*/")
    }

    sql.string <- paste(sql.string, line)
  }

  close(con)
  return(sql.string)
}

'%!in%' <- function(x,y)!('%in%'(x,y))
```

# Loading queries and extracting the data

Loads all queries from the sql files in the extraction folder and runs them into RBigQuey to extract the data.

```{r}
apache_related<-run_query(getSQL("sql/apache_related.sql" ))
hct <- run_query(getSQL("sql/hct.sql" ))
# removes missing values
hct<-hct[complete.cases(hct),]

#Remove chronic AKI, creat, rrt patients
# chronicAKI<-run_query(getSQL('sql/aki/chronicAKI.sql'))
# baseline_creat<-run_query(getSQL('sql/aki/baseline_creat.sql'))
# peakcreat48h<-run_query(getSQL('sql/aki/peakcreat48h.sql'))
# peakcreat7days<-run_query(getSQL('sql/aki/peakcreat7days.sql'))

# renal replacement therapy
# first_rrt <-run_query(getSQL('sql/aki/first_rrt.sql'))

# BMI, weight, age, gender, type of unit that patients was admitted, location from admission, length of stay before admission, height
IO_exclusion <- run_query(getSQL('sql/IO_exclusion.sql'))
charlson_score <- run_query(getSQL('sql/charlson_score.sql'))

# gathers patientunitstayids with reliable fluid data
all_reliable <- run_query (getSQL("sql/all_reliable.sql"))

#demographic
demographic <- run_query(getSQL('sql/demographics.sql'))

# SOFA related variables
sofa_cv_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_cv_day1_to_day4.sql"))
sofa_renal_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_renal_day1_to_day4.sql"))
sofa_respi_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_respi_day1_to_day4.sql"))
sofa_3others_day1_to_day4 <- run_query(getSQL("sql/sofa/sofa_3others_day1_to_day4.sql"))

# List of patients that are bleeding we want to exclude.
patient_inexcluded_icd9 <- run_query(getSQL('sql/patient_inexcluded_icd9.sql'))

# Sepsis patients according to angus criteria
sepsis <- run_query(getSQL('sql/sepsis.sql'))

# Fluid data
fluidIntake <- run_query(getSQL("sql/intake_output.sql"))
```

## SOFA eICU calcuation

```{r}
sofa_total_day1_to_day4<-sqldf('
SELECT selected_cohort.patientunitstayid,

max(sofa_cv_day1_to_day4.sofa_cv_day1 + sofa_respi_day1_to_day4.sofa_respi_day1 + sofa_renal_day1_to_day4.sofarenal_day1 + sofa_3others_day1_to_day4.sofacoag_day1 + sofa_3others_day1_to_day4.sofaliver_day1 + sofa_3others_day1_to_day4.sofacns_day1) AS sofatotal_day1,

max(sofa_cv_day1_to_day4.sofa_cv_day2 + sofa_respi_day1_to_day4.sofa_respi_day2 + sofa_renal_day1_to_day4.sofarenal_day2 + sofa_3others_day1_to_day4.sofacoag_day2 + sofa_3others_day1_to_day4.sofaliver_day2 + sofa_3others_day1_to_day4.sofacns_day2) AS sofatotal_day2,

max(sofa_cv_day1_to_day4.sofa_cv_day3 + sofa_respi_day1_to_day4.sofa_respi_day3 + sofa_renal_day1_to_day4.sofarenal_day3 + sofa_3others_day1_to_day4.sofacoag_day3 + sofa_3others_day1_to_day4.sofaliver_day3 + sofa_3others_day1_to_day4.sofacns_day3) AS sofatotal_day3,

max(sofa_cv_day1_to_day4.sofa_cv_day4 + sofa_respi_day1_to_day4.sofa_respi_day4 + sofa_renal_day1_to_day4.sofarenal_day4 + sofa_3others_day1_to_day4.sofacoag_day4 + sofa_3others_day1_to_day4.sofaliver_day4 + sofa_3others_day1_to_day4.sofacns_day4) AS sofatotal_day4

FROM 
selected_cohort
INNER JOIN sofa_cv_day1_to_day4 ON  selected_cohort.patientunitstayid = sofa_cv_day1_to_day4.patientunitstayid
INNER JOIN sofa_respi_day1_to_day4  ON selected_cohort.patientunitstayid = sofa_renal_day1_to_day4.patientunitstayid
INNER JOIN sofa_renal_day1_to_day4  ON selected_cohort.patientunitstayid = sofa_respi_day1_to_day4.patientunitstayid
INNER JOIN sofa_3others_day1_to_day4  ON selected_cohort.patientunitstayid = sofa_3others_day1_to_day4.patientunitstayid
GROUP BY selected_cohort.patientunitstayid
ORDER BY selected_cohort.patientunitstayid
')
```

## Fluid data

```{r}
# fluidIntake <- run_query(getSQL("sql/intake_output.sql" ))
fluidIntake<-fluidIntake[complete.cases(fluidIntake),]
fluidIntake<-fluidIntake%>%
  mutate( totalFluid = (intakes-outputs))
```

# Merging all datasets

## Patients with sepsis (not bleeding) 

```{r}
# we want to include septic patients that are not bleeding.

print('Septic patients:')
nrow(sepsis)

print('Patients bleeding')
nrow(patient_inexcluded_icd9)

patient_included_vector <- sepsis[sepsis$patientunitstayid %!in% patient_inexcluded_icd9$patientUnitStayID  ,]
patient_included <- data.frame(patientunitstayid=integer(length(patient_included_vector))) 
patient_included$patientunitstayid<-patient_included_vector

print('Septic patients that are not bleeding')
nrow(patient_included)

```


## demographics and too much output

```{r}
selected_cohort<-inner_join(demographic,patient_included)

#exlusion based on output amount
selected_cohort<-sqldf('
SELECT * FROM 
selected_cohort
LEFT JOIN
IO_exclusion
USING
(patientunitstayid)
WHERE
IO_exclusion.patientunitstayid IS NULL
')
```

# Final Join

```{r}
# We are using a left join to join them
vol_leak_index_dataset<-Reduce(function(...) merge(..., all.x=TRUE), list(
   selected_cohort
  ,apache_related
  ,hct
  ,fluidIntake
  # ,AKIlist_final
  ,charlson_score
  ,sofa_total_day1_to_day4
))

nrow(vol_leak_index_dataset)
a<-nrow(vol_leak_index_dataset)

# Ensures that patients have reliable fluid data

vol_leak_index_dataset<-sqldf('
SELECT * FROM
vol_leak_index_dataset
INNER JOIN
all_reliable
USING(patientunitstayid)
')
nrow(vol_leak_index_dataset)
b<-nrow(vol_leak_index_dataset)


a-b
# vol_leak_index_dataset<-sqldf('
# SELECT * FROM
# vol_leak_index_dataset
# INNER JOIN
# all_reliable
# USING(patientunitstayid)
# ')

# Ensures fluid intake is greater than 0.
vol_leak_index_dataset<-sqldf('
SELECT * FROM
vol_leak_index_dataset
WHERE totalFluid > 0 ')
```

## Creating new variables

### Leaking Index

```{r}
vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
    # leaking_index=((mean_hct_24_36hrs/first_hct_6hrs)-1)*body_surface_area*1561
    leaking_index=((mean_hct_24_36hrs - first_hct_6hrs) / totalFluid) * body_surface_area
  )  
```

#### Data Imputation

```{r}
# Median imputation for outliers
#outliers <- boxplot(vol_leak_index_dataset$leaking_index, plot=FALSE)$out
#vol_leak_index_dataset$leaking_index[which(vol_leak_index_dataset$leaking_index %in% outliers),]<-median(vol_leak_index_dataset$leaking_index)

vol_leak_index_dataset<-vol_leak_index_dataset%>%mutate(
  leaking_index=if_else(
  leaking_index %in% outliers,
  median(leaking_index)
  ,leaking_index
  )
)
```


## SOFA Delta

|                   | Sofa day 2 No change | Sofa day 2 Increases | Sofa day 2 Decreases |
|-------------------|----------------------|----------------------|----------------------|
| High Sofa day 1   | Bad                  | Bad                  | Good                 |
| Medium Sofa day 1 | Bad                  | Bad                  | Good                 |
| Low Sofa day 1    | Good                 | Bad                  | Good                 |

0 Means GOOD Oucome, 1 means BAD Outcome

```{r}

vol_leak_index_dataset<-vol_leak_index_dataset%>%
  mutate(
    # leaking_index=((mean_hct_24_36hrs/first_hct_6hrs)-1)*body_surface_area*1561
    leaking_index=((mean_hct_24_36hrs - first_hct_6hrs) / totalFluid) * body_surface_area
    #,delta_sofa=sofatotal_day2-sofatotal_day1
    ,q_leaking_index=as.numeric(cut2(leaking_index, g=4))
    ,bin_leaking_index=if_else(q_leaking_index==4,1,0)
    ,t_sofatotal_day1=as.numeric(cut2(sofatotal_day1, g=3))
    ,delta_sofa=case_when(
      
      t_sofatotal_day1 == 3 & sofatotal_day2 == sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 3 & sofatotal_day2 >  sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 3 & sofatotal_day2 <  sofatotal_day1 ~ 0,
      
      t_sofatotal_day1 == 2 & sofatotal_day2 == sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 2 & sofatotal_day2 >  sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 2 & sofatotal_day2 <  sofatotal_day1 ~ 0,     
      
      t_sofatotal_day1 == 1 & sofatotal_day2 == sofatotal_day1 ~ 0,
      t_sofatotal_day1 == 1 & sofatotal_day2 >  sofatotal_day1 ~ 1,
      t_sofatotal_day1 == 1 & sofatotal_day2 <  sofatotal_day1 ~ 0, 
      
      
    )
                                 
)

vol_leak_index_dataset$q_leaking_index<-as.factor(vol_leak_index_dataset$q_leaking_index)

#vol_leak_index_dataset$actualhospitalmortality<-as.factor(vol_leak_index_dataset$actualhospitalmortality)

# gender mapping
vol_leak_index_dataset<-vol_leak_index_dataset%>% filter(!(gender == ''))
vol_leak_index_dataset$gender[vol_leak_index_dataset$gender=='Male']<-1
vol_leak_index_dataset$gender[vol_leak_index_dataset$gender=='Female']<-2
# vol_leak_index_dataset$gender<-as.factor(vol_leak_index_dataset$gender)


```

## Selecting/Imputing/removing data

```{r}
selected_df <- vol_leak_index_dataset%>%mutate(
  highest_q_leaking_index=if_else(
    q_leaking_index==4,1,0
  ))%>%dplyr::select(
    patientunitstayid
    ,hosp_mortality
    ,age_fixed
    ,gender
    ,final_charlson_score
    ,apachescore
    ,highest_q_leaking_index
    ,delta_sofa
  )
selected_df<-selected_df[complete.cases(selected_df),]

selected_df$highest_q_leaking_index<-as.factor(selected_df$highest_q_leaking_index)
```

# Dataset report

```{r}
vol_leak_index_dataset <- vol_leak_index_dataset[order(vol_leak_index_dataset$patientunitstayid),]
view(dfSummary(selected_df))

sofamissing = sum(is.na(vol_leak_index_dataset$sofatotal_day1)) / length (vol_leak_index_dataset$sofatotal_day1)
bmimissing = sum(is.na(vol_leak_index_dataset$BMI)) / length (vol_leak_index_dataset$BMI)
weightmissing = sum(is.na(vol_leak_index_dataset$weight)) / length (vol_leak_index_dataset$weight)
```

# Export dataset

```{r}
write.csv(vol_leak_index_dataset,'vol_leak_index_dataset.csv')
```

# Train and test datasets creation

## Spliting de-identified data into testing and training, balanced version.

We want the data to be sampled randomly but always the same way and we want to be sure that train and test must be balanced.

```{r}
# Creating id for partition 
selected_df['id']<- seq.int(nrow(selected_df))
## set the seed to make our partition reproducible
set.seed(123)
# createDataPartition: "the random sampling is done within the levels of y when y is a factor in an attempt to balance the class distributions within the splits."
## 75% of the sample size
train_idx <- createDataPartition(as.factor(selected_df$actualhospitalmortality), times = 1, p = 0.75, list=F)

train <- selected_df[train_idx, ]
test <- selected_df[-train_idx, ]

#Checking outcome is actually balanced
round(prop.table(table(train$actualhospitalmortality)),2)
round(prop.table(table(test$actualhospitalmortality)),2)
```

## Separating datasets into outcome and exposures

```{r}
# train dataset
train_X<-train[, names(train)!= "actualhospitalmortality"]
train_Y<-train$actualhospitalmortality
  
# test dataset
test_X<-test[, names(test)!= "actualhospitalmortality"]
test_Y<-test$actualhospitalmortality 
```







