---
title: "02_model_training"
author: "Miguel Ángel Armengol"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_notebook:
    code_folding: hide
    number_sections: yes
    theme: flatly
    toc: yes
    toc_float: yes

knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = paste0(substr(inputFile,1,nchar(inputFile)-4)," ",Sys.Date(),'.html')) })
---


# Environment

```{r message=FALSE, warning=FALSE}
require(foreign)
require(ggplot2)
require(MASS)
require(Hmisc)
require(reshape2)
require(caret)
require(boot)
require(pROC)
library(mlbench)
library(MLmetrics)
library(plotly)
library(gbm)
library(xgboost)
library(oddsratio)
library(ggplot2)
library(survey)
library(table1)
library(sandwich)
```

# ML: Random Hyperparameter Tunning

The default method for optimizing tuning parameters in train is to use a grid search. This approach is usually effective but, in cases when there are many tuning parameters, it can be inefficient. An alternative is to use a combination of grid search and racing. Another is to use a random selection of tuning parameter combinations to cover the parameter space to a lesser extent.

Using [caret](https://topepo.github.io/caret/).

_We can adress later the tuning parameters approach_

```{r}
fitControl <- trainControl(method = "repeatedcv",
                           number = 10,
                           repeats = 10,
                           classProbs = TRUE,
                           summaryFunction = twoClassSummary,
                           search = "random")
```

### Random selection of tuning parameter combinations

Here we are first adressing several Machine Learning methods.
There are more methods that can be addressed [Available Models in caret::train](https://rdrr.io/cran/caret/man/models.html)

```{r}
gbmFit <- train( hosp_mortality ~ 
                + apachescore
                + final_charlson_score
                + q_leaking_index
                ,data = train,
                method = "gbm",
                trControl = fitControl,
                verbose = FALSE,
                metric = "ROC" ## Specify which metric to optimize
)

svmFit <- train( hosp_mortality ~
                + apachescore
                + final_charlson_score
                + q_leaking_index
                ,data = train,
                method = "svmRadial",
                trControl = fitControl,
                preProc = c("center", "scale"),
                tuneLength = 8,
                metric = "ROC" ## Specify which metric to optimize
)

rfFit <- train( hosp_mortality ~
                + apachescore
                + final_charlson_score
                + q_leaking_index
                ,data = train,
                method = "rf",
                trControl = fitControl,
                verbose = FALSE,
                metric = "ROC" ## Specify which metric to optimize
)

xgbFit <- train( hosp_mortality ~
                + apachescore
                + final_charlson_score
                + q_leaking_index
                ,data = train,
                method = "xgbTree",
                trControl = fitControl,
                verbose = FALSE,
                metric = "ROC" ## Specify which metric to optimize
)

nnFit <- train( hosp_mortality ~
                + apachescore
                + final_charlson_score
                + q_leaking_index
                ,data = train,
                method = "nnet",
                trControl = fitControl,
                verbose = FALSE,
                metric = "ROC" ## Specify which metric to optimize
)


lrFit <- train( hosp_mortality ~
                + apachescore
                + final_charlson_score
                + q_leaking_index
                ,data = train,
                method = "LogitBoost",
                trControl = fitControl,
                verbose = FALSE,
                metric = "ROC" ## Specify which metric to optimize
)

gamFit <- train( hosp_mortality ~ 
                + apachescore
                + final_charlson_score
                + q_leaking_index
                ,data = train,
                method = "gam",
                trControl = fitControl,
                verbose = T,
                metric = "ROC" ## Specify which metric to optimize
)
```

### Best models comprarision

```{r}
resamps <- resamples(list( gbmFit = gbmFit
                          ,svmFit = svmFit
                          ,rfFit  = rfFit
                          ,xgbFit = xgbFit
                          ,nnFit = nnFit
                          ,lrFit = lrFit
                          ,gamFit = gamFit
                          ))
summary_resamps<-summary(resamps)

summary_resamps<-as.data.frame(summary_resamps$statistics)
summary_resamps
```


## Selecting the model with the best performance

```{r}
# we save the best performing model (based on its ROC) and its name
best_performing_model<-get(
  rownames(summary_resamps[which(summary_resamps$ROC.Median==max(summary_resamps$ROC.Median))]
)
)
#manually select it
best_performing_model<-gamFit
best_performing_model_name<-best_performing_model$method # extracts name as string from model
```

We can see **`r best_performing_model_name`** is the model with the best performance, with a Median AUROC of **`r max(summary_resamps$ROC.Median)`**.  

Its best Random Hyperparameter Tune was:  
`r best_performing_model$bestTune`

## Evaluating the predictor on our test dataset

### Creating prediction-probabilities dataset

```{r}
prediction_probabilities<-predict(best_performing_model, newdata = test,type = "prob") # We create the probabilities dataset using our best performing model.

final_predictions<-cbind(test_Y,prediction_probabilities) # we bind our prediction with the actual data
final_predictions<-rename(final_predictions, obs = test_Y) # the function twoClassSummary reads the actual outcome as 'obs'
final_predictions['pred']<-ifelse(final_predictions$ALIVE > .83 # we have set the threshold in .5 this can be optimized until best performance is achieved
                                  , 'ALIVE','EXPIRED'
)

# Setting proper data types
final_predictions$obs<-as.factor(final_predictions$obs)
final_predictions$pred<-as.factor(final_predictions$pred)
```

### Geting evaluation insights

```{r}
insights_1<-as.data.frame(twoClassSummary(final_predictions, lev = levels(final_predictions$obs)))
names(insights_1)<-best_performing_model_name
insights_1<-t(insights_1) # we traspose it for better merging it.

insights_2<-as.data.frame(prSummary(final_predictions, lev = levels(final_predictions$obs)))
names(insights_2)<-best_performing_model_name
insights_2<-t(insights_2) # we traspose it for better merging it.

evaluation_insights<-cbind(insights_1,insights_2)
evaluation_insights<-as.data.frame(evaluation_insights)
evaluation_insights<-round(evaluation_insights,2)
evaluation_insights$Recall <-NULL # We already have specificity which is = recall
evaluation_insights$AUC<-NULL # We select the ROC from the first package so we remove this parameter
#renaming metric
evaluation_insights<-evaluation_insights%>%rename(AUROC = ROC) 

# we traspose the data for its representation
evaluation_insights<-t(evaluation_insights)
evaluation_insights<-as.data.frame(evaluation_insights)
evaluation_insights['Metrics']<-rownames(evaluation_insights)

# how to order the bars
evaluation_insights$Insights <- factor(evaluation_insights$Metrics
                     , levels = unique(evaluation_insights$Insights)[order(evaluation_insights$Metrics, decreasing = T)])

p <- plot_ly(
  data = evaluation_insights,
  x = evaluation_insights[,1],
  y = ~Metrics,
  text = evaluation_insights[,1],
  textposition='auto',
  type = "bar") %>%
  layout( title = paste(best_performing_model_name,"Model Metrics"))

p
```

## Variables Importance

```{r message=TRUE, warning=TRUE}
ggthemr('flat')
varimp<-ggplot(varImp(best_performing_model, scale = T))+theme_minimal() 
ggplotly(varimp)
```

## Odds ratio 

Valid for gam logreg and glm.

We are using the or_gam function

```{r}
# Odds Ratio Calculations

train_for_or_calculation<-train
# or_gam only deals with numerical vars
train_for_or_calculation$hosp_mortality[train_for_or_calculation$hosp_mortality=='ALIVE']<-0 
train_for_or_calculation$hosp_mortality[train_for_or_calculation$hosp_mortality=='EXPIRED']<-1
train_for_or_calculation$hosp_mortality<-as.numeric(train_for_or_calculation$hosp_mortality)

# We need to run the gam separately (the function I am using is not compatible with th gam object caret creates, so I am training a new model just using gam().

gam_for_or_calculation<-gam(hosp_mortality ~ 
                + apachescore
                + final_charlson_score
                + q_leaking_index
                ,data = train_for_or_calculation)

# every exposure needs to be addressed separately the min value as a reference.
# every factor in the categorical exposures need to be addressed separately taking into account the min value as a reference.

# q_leaking_index Ors
or_gam(data = train_for_or_calculation, model = gam_for_or_calculation, pred = "q_leaking_index",values = c("1", "2"))
or_gam(data = train_for_or_calculation, model = gam_for_or_calculation, pred = "q_leaking_index",values = c("1", "3"))
or_gam(data = train_for_or_calculation, model = gam_for_or_calculation, pred = "q_leaking_index",values = c("1", "4"))

# rest of the exposures

or_gam(data = train_for_or_calculation, model = gam_for_or_calculation, pred = "apachescore",values = c(min(train_for_or_calculation$apachescore), max(train_for_or_calculation$apachescore)))
or_gam(data = train_for_or_calculation, model = gam_for_or_calculation, pred = "final_charlson_score",values = c(min(train_for_or_calculation$final_charlson_score), max(train_for_or_calculation$final_charlson_score)))
```

# Causal inference

## Propensity Score Matching - logistic regression

```{r}
#first of all we create a copy of the dataset
vli_prop_score_dataset<-selected_df
# Fit a propensity score model: logistic regression
# outcome is the treatment here.
ps_model <- glm(highest_q_leaking_index ~ 
                  apachescore
                + age_fixed
                + gender
                + final_charlson_score
                + totalFluid
                , family=binomial()
                , data=vli_prop_score_dataset)

summary(ps_model)

OR_table<-as.data.frame(round(exp(cbind(OR=coef(ps_model), confint.default(ps_model))),2))
options(scipen=999) # disable scientific notation
OR_table

# Value of propensity score for each patient

pscore<-ps_model$fitted.values

treatment_pscore<-cbind(vli_prop_score_dataset%>%dplyr::select(highest_q_leaking_index),pscore)

ggplot(treatment_pscore, aes(x=pscore, fill = highest_q_leaking_index)) + 
  geom_histogram(alpha = 0.3,position="identity")+scale_fill_manual(name="Treatment: Prob of Highest Q of VLI",values=c("#1abc9c","#f1c40f","#2980b9","#e74c3c"),labels=c("Q1","The rest"))+theme_minimal()
```


## Inverse probability weighting

```{r}
# create weight
hr_prop_score_dataset<-vli_prop_score_dataset%>%mutate(
  weight=if_else(highest_q_leaking_index==1 # TODO: what is the control group? People in q1,q2,q3
                 ,1/(pscore)
                 ,1/(1-pscore))
)

hr_prop_score_dataset$weight<-as.numeric(as.character(hr_prop_score_dataset$weight))

ggplot(hr_prop_score_dataset, aes(x = weight, fill = highest_q_leaking_index)) +
   geom_density(alpha = 0.5, colour = "grey50") +
   geom_rug() +
   scale_x_log10(breaks = c(1, 5, 10, 20, 40)) +   ggtitle("Distribution of inverse probability weights")+ scale_fill_manual(name="Treatment: Highest Quartile of VLI",values=c("#1abc9c","#f1c40f","#2980b9","#e74c3c"),labels=c("Not on Q4","On Q4"))+theme_minimal()

# apply weights to data
weighted_data<-svydesign(ids = ~ patientunitstayid, data = hr_prop_score_dataset,
                         weights = ~weight
                         )

# weighted table 1

weighedtable<- svyCreateTableOne( vars= c("age_fixed", "gender","final_charlson_score","apachescore"), strata = "highest_q_leaking_index", data = weighted_data, test = F)

# weighted table 1 wih standarized mean differences (SMD)
as.data.frame(print(weighedtable,smd=T))
# Ignore SD and sample sizes, ONLY SMD is reliable

```

In tableone: Ignore SD and sample sizes, ONLY SMD is reliable


## Marginal structural modeling

### Hospital Mortality prediction

#### Relative risk with CI

```{r warning=FALSE}
# Obtaining causal relative risk. Weighed GLM

glm_model_rr<-glm(
  as.integer(hosp_mortality) ~ 
  as.factor(highest_q_leaking_index)  # treatment is the only exposure now
, weights = weight  
, family = binomial(link = log) # we are using the log link since we are interested in the relative risk
, data= hr_prop_score_dataset
)

# summary of the glm_model_rr final model
# beta Inverse probability weighting
betaiptw<-coef(glm_model_rr)

# to properly account for weighting, we are going to use asymptotic (sandwich) variance

SE<-sqrt(diag(vcovHC(glm_model_rr,type = "HC0"))) # getting the standard error.

# we get point estimate and CI for relative risk

beta_causal_relative_risk<-exp(betaiptw[2]) # we need to exponientiate since we logged before
lCI_q1<-exp(betaiptw[2]-1.96*SE[2])
uCI_q1<-exp(betaiptw[2]+1.96*SE[2])

beta_final<-as.data.frame(cbind(beta_causal_relative_risk, lCI_q1, uCI_q1
                    ))

rownames(beta_final)<-''

beta_final
```

#### Risk difference with CI

```{r warning=FALSE}
# Obtaining Obtaining risk difference with CI

glm_model_diff<-glm(
  as.integer(hosp_mortality) ~ 
  as.factor(highest_q_leaking_index) # treatment is the only exposure now
, weights = weight  
, family = binomial(link = 'identity') # we are using the identity link since we are interested in risk difference
, data= hr_prop_score_dataset
)

# summary of the glm_model_diff final model
# beta Inverse probability weighting
betaiptw<-coef(glm_model_diff)

# to properly account for weighting, we are going to use asymptotic (sandwich) variance

SE<-sqrt(diag(vcovHC(glm_model_diff,type = "HC0"))) # getting the standard error.

# we get point estimate and CI for relative risk

beta_risk_diference<-exp(betaiptw[2]) # we need to exponientiate since we logged before
lCI_q1<-exp(betaiptw[2]-1.96*SE[2])
uCI_q1<-exp(betaiptw[2]+1.96*SE[2])

beta_final<-as.data.frame(cbind(beta_risk_diference, lCI_q1, uCI_q1
                    ))

rownames(beta_final)<-''

beta_final
```


### Q Delta Sofa prediction

#### Relative risk with CI

```{r warning=FALSE}
# Obtaining causal relative risk. Weighed GLM

glm_model_rr<-glm(
  delta_sofa ~ 
  as.factor(highest_q_leaking_index)  # treatment is the only exposure now
, weights = weight  
, family = poisson(link = log) # we are using the log link since we are interested in the relative risk
, data= hr_prop_score_dataset
)

# summary of the glm_model_rr final model
# beta Inverse probability weighting
betaiptw<-coef(glm_model_rr)

# to properly account for weighting, we are going to use asymptotic (sandwich) variance

SE<-sqrt(diag(vcovHC(glm_model_rr,type = "HC0"))) # getting the standard error.

# we get point estimate and CI for relative risk

beta_causal_relative_risk<-exp(betaiptw[2]) # we need to exponientiate since we logged before
lCI_q1<-exp(betaiptw[2]-1.96*SE[2])
uCI_q1<-exp(betaiptw[2]+1.96*SE[2])

beta_final<-as.data.frame(cbind(beta_causal_relative_risk, lCI_q1, uCI_q1
                    ))

rownames(beta_final)<-''

beta_final
```

#### Risk difference with CI

```{r warning=FALSE}
# Obtaining Obtaining risk difference with CI

glm_model_diff<-glm(
  delta_sofa ~ 
  as.factor(highest_q_leaking_index) # treatment is the only exposure now
, weights = weight  
, family = poisson(link = 'identity') # we are using the identity link since we are interested in risk difference
, data= hr_prop_score_dataset
)

# summary of the glm_model_diff final model
# beta Inverse probability weighting
betaiptw<-coef(glm_model_diff)

# to properly account for weighting, we are going to use asymptotic (sandwich) variance

SE<-sqrt(diag(vcovHC(glm_model_diff,type = "HC0"))) # getting the standard error.

# we get point estimate and CI for relative risk

beta_risk_diference<-exp(betaiptw[2]) # we need to exponientiate since we logged before
lCI_q1<-exp(betaiptw[2]-1.96*SE[2])
uCI_q1<-exp(betaiptw[2]+1.96*SE[2])

beta_final<-as.data.frame(cbind(beta_risk_diference, lCI_q1, uCI_q1
                    ))

rownames(beta_final)<-''

beta_final
```

